Discussion Draft

### 2020-04-04 doc/Discussion.rmd
more small changes

The common direction of heterochiasmy is female biased, however male biased heterochiasmy examples are not especially rare. Our results are novel from previous examples of male biased heterochiasmy species because the panel of strains have much shorter evolutionary distances. This context gives us insights to sex-specific evolution of recombination rates.

The prediction from two of the modifier models, haploid selection and SACE, state that males will generally evolve lower recombination rates. Male biased heterochiasmy is predicted under the two locus modifier model when XXXX.  <does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).>

There are no prediction for male biased heterochiasmy evolution direction under the COM <pairing model>, but it should be noted that this model was not designed to incorporate this type of data. This model was design to explain the positive interference pattern for of the recombination landscape by incorporating a cell mechanics of rapid prophase movements (RPM) that are thought to facilitate homolog pairing in many species.

Under the spindle based model faster male evolution in male genome wide recombination rates is primarily driven the strength of the spindle assembly checkpoint (SAC) <acting on the chromosome structure>.

The genes and pathways are will known, we refer the reader to [@laneKauppi2019].

@@

Unstable bivalents are detected on the MI spindle, the SAC delays the cell cycle which potentially allows time for lose bivalents to correctly attach to the MT spindle. If the 'unattached' persists apoptosis will be triggered. Genetic variants which increase the number of unstable bivalents, will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Thus the SAC can lead to stabilizing selection on the chromosome structure.

@@

Interference is a fundamental aspect of the recombination landscape. -- genome-wide recombination rate and chromatin compaction

Some basics of the constraints bounds of interference are known, yet there are still many unknowns regarding the relationship of genome wide recombination rate and interference strength.

@@

# REVIEW

We propose that the spindle based selection model is the most parsimounious with our results and known recombination patterns from the literautre.

  - It matches fine scale differences in how DSB hotspots had different sex specific strengths with varied due to sex, epigenetc landscape and the gene architecture (Brick et al)

 - incorporates know cellular / molecular mechanisms with a very common pattern sexual dimorphic recombination rate landscape.

- extends modifer model to explain a chromosome based / level pattern
 _weaknesses_, and ways the model diverges from our results

 - the simulation present in the paper are too weak to support the rapid spread of such a modifer system (selection coefficients are too weak to support the evolution -- spread of such a modifer.)
- Other ways this model departs from our results, 
-Always predicting male will be lower recombination rates (or risk breaking up the epistatsis blocks) -- which doesn't fit with our results of faster male specific evolution in gwRR.
However / while not explicitaly arameterized or predicted in this model,
 -- this model predicts larger chuncks of chromosomes will segregate together
 (as outlined in Veller et al)
our results of stronger male interference and stronger interference in high rec strains have the same consequences as predited from this modle

```{r final.review.outline, echo=FALSE}
# comes down to explaning the interference correlation

# spindel selection -- 
#good for sex differences because it frames heterochiasmy within the context of meiosis

#can fit the interference evolution due to the (pressumed) effect on bivalent shape

#caveats and unanswered parts

# SACE is runner up

@@

Be aware of sex-specific and sex average biases. Some cytlogy models (Celegans) mostly focus on oogenesis  and genetic maps are often from hermaphrodites. (lack of data from one sex). Recognize the value of Chromosome level measures and incorporating chromosome behavior into models.

Borader communication across sub-fields of cell physiology and biophysics for understanding recombination rate evolution. Utilize mitosis systems given the conservation of the molecular pathway. THink of meiosis as a program.

Dark matter of the genome. We can't really explude that some of the recombination rate variation is due to strucutreal variation (more of less genome). (since everything is mapped to B6). (also maybe there are crossovers in the centromeres and telomeres -- which might escape detected in linkage maps)

@@

How does elevated rates of precocious sister centromere seperation effect oogenesis?

@@

#blah

The SAC 'read's that kinetochores (the movement machinery built on top of the centromere) are connected to the MT of the spindle originating from the correct pole of the cell. (tension is )

centromeres have single and opposite pole connection (synthellic?) and that the sister cohesion connecting the homologs -- sets up the tension -- opposes the pull from each of the poles.

Efficient normal SAC -- works by pausing entry into anaphase or triggers apoptosis -- if a unpaired / tension-lacking tetrad is detected
(sperm SAC is finer tuned (can detect 1-2 tetrads lacking tension -- ), while the female SAC  (has a higher threshold for being triggered  (4-5?)
(Land Kauppi,

(relaxed selection - / inefficient SAC -- in females -- fits with higher rate of achiasmata bivalents (there might be alternative mechanisms facilitating achiasmate segregation)

  Mechanisms for sex differences in the strength of the SAC are the cell volumne (diffusion of signal molecules) and the centrosome being present at the spindle have been proposed in the literature.
(testing / proving the mechanism for the new optima is for the high recombining group in males is out of the scope of this paper, but we propose some hypothesis are suggested below.
 but the evolutionary pattern fits that of directional.

>>>>

The common direction of heterochiasmy is female biased, however male biased heterochiasmy examples are not especially rare. Our results are novel from previous examples of male biased heterochiasmy species because the panel of strains have much shorter evolutionary distances. This context gives us insights to sex-specific evolution of recombination rates.  
Male biased heterochiasmy is predicted under the two locus modifier model when XXXX.  <does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).>

**strict SAC drives directional selection**,  
**genetic variants which lead to 'unstable chromosome structure will be selected agaisnt**  
**sex differences in SAC strength are connected to conserved differences in gametogensis**

Under the **spindle based** model faster male evolution in male genome wide recombination rates is primarily driven the strength of the spindle assembly checkpoint (SAC) acting on the chromosome structure at metaphase alignment. We will not go into much detail on the mechanism of the SAC, The genes and pathways are will known, we refer the reader to [@laneKauppi2019].

@@

Unstable bivalents are detected on the MI spindle, the SAC delays the cell cycle which potentially allows time for lose bivalents to correctly attach to the MT spindle. If the 'unattached' persists apoptosis will be triggered. Genetic variants which increase the number of unstable bivalents, will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle.

@@

Interference is a fundamental aspect genome-wide recombination rate the, chromatin compaction and the resulting recombination landscape.

@@

@@

```{r future.step.plan, echo=FALSE}
#still preliminary
```

- sex-specific and sex average biases. Some cytology models (Celegans) mostly focus on oogenesis  and genetic maps are often from hermaphrodites. (lack of data from one sex). Recognize the value of Chromosome level measures and incorporating chromosome behavior into models.

Broader communication across sub-fields of cell physiology and biophysics for understanding recombination rate evolution. Utilize mitosis systems given the conservation of the molecular pathway. Think of meiosis as a program.

Dark matter of the genome. We can't really exclude that some of the recombination rate variation is due to structural variation (more of less genome). (since everything is mapped to B6). (also maybe there are crossovers in the centromeres and telomeres -- which might escape detected in linkage maps)

@@

How does elevated rates of precocious sister centromere separation effect oogenesis?

@@

- chromatin interference
